What is an Ancestral Recombination Graph?

### 2022-01-31 paper.tex
Added section on phylogenetic networks; edited section on parsimonious ARG reconstruction

Recent breakthroughs large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2021unified}
and data representation~\citep{kelleher2016efficient}
have raised the genuine prospect of genealogical analysis becoming a standard part
of the population and statistical genetics toolkit.
Applications using these inferred ancestries as input have
begun to appear~\citep{osmond2021estimating} and many more are sure to
follow~\citep{harris2019database}. This vibrant research area, however,
has a significant difficulty, which, if not addressed, will substantially hamper
progress. We currently lack a well-defined data-model and shared terminology
to discuss recombining genealogies,
leading to basic errors in statements about what different inference methods
produce as output (see the XXX section below), as well as presenting serious
difficulties in either comparing the outputs of different methods or
(from a users perspective) using multiple inference methods in an analysis.

@@

% NOTE: it would probably be helpful to have a drawing of the ARG process
% in-progress here to help us talk about it. What is the full state
% that is remembered as we go through the process? It does help to think
% about the process as choosing edges to merge, not nodes.
The Ancestral Recombination Graph (ARG) was later introduced
Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,
griffiths1997ancestral}, and is a
closely related and complementary treatment of the same process. Where
Hudson's algorithm is focused on the efficient \emph{simulation} of the
process, the original ARG literature is focused on mathematical
results about the stochastic process. Griffiths and colleagues formulated the
ARG as a branching-coalescing stochastic process. We imagine a graph in which
each edge corresponds to an extant lineage and nodes are events in the
process (the initial $n$ leaf nodes are ``sampling'' events). As in Hudson's
algorithm, common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events are different
to Hudson's algorithm, however. Rather than occuring at a rate which depends on
the specific distribution of ancestral material among lineages, the
rate of recombination is simply $k \rho (m - 1)$ in the original ARG
formulation. At a recombination event we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < L$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMCRA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. The graph structure and breakpoints associated with with
recombination nodes provides sufficient information to later recover the marginal
trees (see the next section for more details and discussion on this point).

@@

The first method to explicitly produce an ARG as output is SHRUB developed
by \citet{song2005efficient}.

@@

\section*{ARG inference methods}
Using the terminology developed here to classify ARGs, we review methods developed
to infer ARGs. We focus our attention on some recent methods and discuss the
properties of the ARGs inferred.

The ARG has been used as a model for recombining DNA sequence data essentially
since it was first formulated. Early methods focused on finding parsimonious
ARGs consistent with a given set of
sequences \citep{hein1990reconstructing, lyngso2005minimum}.
However, identifying the most parsimonious ARG for a given sequence data set
is NP-hard \citep{wang2001perfect}, and subsequent methods have resorted to
various heuristics \citep{fallon2013acg, hein1993heuristic, ignatieva2021kwarg,
kelleher2019inferring, minichiello2006mapping, mirzaei2017rent,
parida2008estimating, song2005efficient, thao2019hybrid, zhang2021biobank}.
The scalability of these methods varies substantially based on the heuristic in use
and the underlying data structure, but the most efficient methods can be practically
used on megabase scale data and tens of thousands of samples under human-like
parameters \citep{kelleher2019inferring, zhang2021biobank}.

\citep{camara2016inference} is another type of parsimony method.

An alternative approach is to treat the ancestry as a latent parameter to be averaged out
by Monte Carlo methods, based either on importance sampling
\citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference}
or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian}.
These methods invariably operated on a representation of the ``little ARG", typically
generated until the grand MRCA. They are extremely computationally expensive,
and applicable to at most hundreds of samples consisting of tens of kilobases with
human-like parameters.

>>>>

Recent breakthroughs in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2021unified}
and data representation~\citep{kelleher2016efficient}
have raised the genuine prospect of genealogical analysis becoming a standard part
of the population and statistical genetics toolkit.
Applications using these inferred ancestries as input have
begun to appear~\citep{osmond2021estimating} and many more are sure to
follow~\citep{harris2019database}. This vibrant research area, however,
has a significant difficulty, which, if not addressed, will substantially hamper
progress. We currently lack a well-defined data-model and shared terminology
to discuss recombining genealogies,
leading to basic errors in statements about what different inference methods
produce as output (see the XXX section below), as well as presenting serious
difficulties in either comparing the outputs of different methods or
(from a users perspective) using multiple inference methods in an analysis.

@@

% NOTE: it would probably be helpful to have a drawing of the ARG process
% in-progress here to help us talk about it. What is the full state
% that is remembered as we go through the process? It does help to think
% about the process as choosing edges to merge, not nodes.
The Ancestral Recombination Graph (ARG) was later introduced by Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,
griffiths1997ancestral}, and is a
closely related and complementary treatment of the same process. Where
Hudson's algorithm is focused on the efficient \emph{simulation} of the
process, the original ARG literature is focused on mathematical
results about the stochastic process. Griffiths and colleagues formulated the
ARG as a branching-coalescing stochastic process. We imagine a graph in which
each edge corresponds to an extant lineage and nodes are events in the
process (the initial $n$ leaf nodes are ``sampling'' events). As in Hudson's
algorithm, common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events are different
to Hudson's algorithm, however. Rather than occuring at a rate which depends on
the specific distribution of ancestral material among lineages, the
rate of recombination is simply $k \rho (m - 1)$ in the original ARG
formulation. At a recombination event we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < L$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMCRA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. The graph structure and breakpoints associated with with
recombination nodes provides sufficient information to later recover the marginal
trees (see the next section for more details and discussion on this point).

@@

The first method to explicitly produce an ARG as output is SHRUB developed
by \citet{song2005efficient}.
% AI: I think \citep{song2003parsimonious} came before.

@@

\section*{Relationship with other types of phylogenetic networks}

The term \emph{phylogenetic network} has been broadly accepted to describe the most general 
type of genealogical graph, defined by \citet{huson2010phylogenetic} as ``any graph used to 
represent evolutionary relationships (either abstractly or explicitly) between a set of taxa 
that labels some of its nodes" (although more narrow and context-dependent definitions can also 
be found in the literature). Explicit rooted phylogenetic networks are rooted DAGs,
with leaves labelled by the taxa (species, groups, individuals or sequences), and nodes classed 
as either reticulation nodes (with in-degree $\geq 2$) or tree nodes.

Some sub-classes of explicit rooted phylogenetic networks are defined based on the evolutionary event that the
reticulation nodes capture \citep{huson2010phylogenetic}. In \emph{reassortment} networks, used to 
model the evolution of viruses, reticulation nodes correspond to exchange of 
(non-recombining) genomic segments between viral particles inside a co-infected host. 
In \emph{hybridisation} networks, reticulation nodes correspond to genetic material from two
different species combining to create a hybrid. In \emph{recombination} networks, the reticulation nodes 
correspond to recombination events (with leaves labelled by sequences and edges labelled by mutations). 
The ARG is a subtype of a recombination network under this definition (restricting to one crossover 
recombination per reticulation node).

Sub-classes of explicit rooted phylogenetic networks can also be defined by restrictions on their topology. 
In \emph{tree-child networks}, every internal node must have at least one child that is a tree node; 
\emph{tree-sibling networks} are constrained by every reticulation node having at least one sibling that is a 
tree node \citep{cardona2008extended}. Further, sub-types can be defined by restrictions on 
the topology of subgraphs that contain the reticulation loops (cycles containing a reticulation node). 
A reticulation loop that does not share any node with another reticulation loop is termed a gall 
\citep[][p.\ 237]{gusfield2014recombinatorics}; a reticulation network containing only galls is 
called a \emph{galled tree}. Reticulation loops that are joined together in a reticulation network 
by at least one shared node form a blob; a \emph{level-$k$ network} is one where the maximal number 
of reticulation loops inside a blob is $k$ \citep{choy2005computing}. Galled trees are level-1 networks, 
and the tractability afforded by their relatively simple structure results in a number of attractive 
combinatorial properties \citep{wang2001perfect, gusfield2004optimal}. We consider the general case of 
an ARG, without applying any of these restrictions.

% AI: have not mentioned any un-rooted networks (split, median, haplotype, reticulograms, etc).

\section*{ARG inference methods}
Using the terminology developed here to classify ARGs, we review methods developed
to \emph{infer} ARGs from sequencing data. We focus our attention on some recent methods and discuss the
properties of the ARGs inferred.

The problem of reconstructing ARGs for samples of recombining sequences has been of significant interest
essentially since the ARG was first defined. Early methods focussed on finding the most 
\emph{parsimonious} ARGs, i.e.\ those minimising the number of 
posited recombination events in the history of the sample \citep{hein1990reconstructing, song2003parsimonious,
lyngso2005minimum}. However, reconstructing a most parsimonious ARG for a given data set
is NP-hard \citep{wang2001perfect}, and in order to allow for larger input datasets,
subsequent parsimony-based methods have resorted to various relaxations and heuristics \citep{hein1993heuristic, 
song2005efficient, wu2008association, camara2016inference, thao2019hybrid, ignatieva2021kwarg}.
The scalability of these methods varies substantially based on the heuristic in use
and the underlying data structure, with the most efficient tools limited to analysing hundreds of sequences.
A number of heuristic methods have also been developed which do not (explicitly) focus on reconstructing
the most parsimonious ARGs, but rather aim for computational efficiency
while seeking to achieve good accuracy of the inferred genealogies
\citep{minichiello2006mapping, parida2008estimating, wu2011new, mirzaei2017rent}; some of these methods \citep{kelleher2019inferring, speidel2019method, zhang2021biobank} can be practically
used on megabase scale data and tens to hundreds of thousands of samples under human-like
parameters.
% AI: just to note, these papers do not all agree in their definition of an ARG. 
% I guess this is one of the points of the present paper, but should we mention this fact?
% Should we also mention that most of these infer just the topology, but some also the times?

An alternative approach is to treat the ancestry as a latent parameter to be averaged out
by Monte Carlo methods, based either on importance sampling
\citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference}
or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian, fallon2013acg}.
These methods invariably operated on a representation of the ``little ARG", typically
generated until the grand MRCA. They are extremely computationally expensive,
and applicable to at most hundreds of samples consisting of tens of kilobases with
human-like parameters.
